A Plasma Torus Around a Young Low-Mass Star

### 2025-06-09 paper/ms.tex
update

% upon AAS submission
\documentclass[11pt,twocolumn,tighten,linenumbers]{aastex7}
%\documentclass[11pt,twocolumn,tighten,linenumbers,trackchanges]{aastex7}
% drafting / arxiv
%\documentclass[11pt,twocolumn,modern]{aastex7}
%\turnoffedit

@@

\received{23 April 2025}
\revised{--}
\accepted{--}
\shorttitle{CPV Plasma Torus}

@@

\begin{figure*}[!t]
  \centering
  \includegraphics[width=0.925\textwidth]{f2.pdf}
  \vspace{-0.2cm}
  \caption{{\bf Photometric evolution of TIC\,141146667 during the Keck/HIRES
  observation (green bar).}  {\bf a,} TESS simple aperture photometry.
  Data gaps were caused by stray light from the Earth (BTJD 3356-3358.5) and
  Moon (BTJD 3359.5-3360.5).  Raw two minute data are in gray; black
  time-averages to ten minute sampling.  {\bf b-c,} Folded TESS light
  curve before and after spectroscopy.  Black now phase-averages to
  100 points per 3.93\ hour cycle.  During BTJD 3352-3356, a state
  switch occurred near BTJD 3353, and the dip at $\phi$$\approx$0.8
  \added{became less pronounced}.  While the \added{dip spanning
  $\phi$=0.6-1.1} was present both before and after the HIRES
  sequence, its photometric shape evolved during the data gap.  The
  orange bar denotes times of spectroscopic transits for the inner two
  H$\alpha$ clumps observed with HIRES (see Figure~\ref{fig:spec}).}
  \label{fig:fulllc}
\end{figure*}

\begin{figure*}[!t]
  \centering
  \includegraphics[width=0.99\textwidth]{f3.pdf}
  \vspace{-0.3cm}
  \captionsetup{labelformat=moviefmt,labelsep=colon}
	\caption{\textbf{Emission from plasma clumps orbiting
TIC\,141146667.}
  The online movie,
  \href{https://lgbouma.com/movies/TIC141146667_sixpanel.mp4}{{\bf
  available here}},
  shows the spectral evolution over five hours.
  {\bf a,} TESS light curve from 5 February 2024 to 26
  February 2024 folded on the 3.93\ hour period.  Black
  points are phase-averaged; gray are the raw data.
  {\bf b,} Keck/HIRES H$\alpha$ spectra from 17 February 2024.
  The continuum is set to unity, and the darkest color is at twice the
  continuum to accentuate emission outside the line core ($|v/v_{\rm
  eq}|>1$, for the equatorial velocity $v_{\rm eq}$=130\,\kms).  While
  emission within the line core comes from the star's chromosphere, the
  sinusoidal emission features are most readily described by a warped
  plasma torus.
  {\bf c,} Individual epochs of Panel b, visible in the online movie.
  The dotted line shows the time-averaged spectrum, $f_{\langle t
  \rangle}$.
  {\bf d,} As in Panel a, but overplotting the
  median-normalized H$\alpha$ light curve at $|v/v_{\rm eq}|<1$.
  {\bf e,} As in Panel b, after subtracting $f_{\langle t
  \rangle}$.  The line core shows H$\alpha$ excesses and decrements
  advancing from the blue to red wings.
  The asymmetric color stretch mirrors the dynamic range of the data.
  {\bf f,} Individual epochs of Panel e, visible in the online
  movie.  \added{A separate version with best-fit sinusoids overplotted is also
  \href{https://lgbouma.com/movies/TIC141146667_sixpanel_sinusoids.mp4}{{\bf
  available}}.}}
  \label{fig:spec}
\end{figure*}

@@

The other CPV analog is the $\sigma$~Ori~E variables, a rare subset of
B stars with radiatively-driven winds that accumulate into warped
plasma tori \citep{Townsend2005}.  These tori tend to have dense
antipodal accumulations of plasma sculpted by tilted-dipole magnetic
fields; these clumps produce broadband optical variability through
bound-free scattering \citep{Townsend2005} and Thomson scattering
\citep{Berry2022}.  For $\sigma$~Ori~E and most of its analogs, the
result is simple light curves that resemble eclipsing binaries, and
time-dynamic H$\alpha$ spectra similar to Figure~\ref{fig:spec}.  The
two known exceptions, HD~37776 and HD~64740, have spectropolarimetric
magnetic field maps indicating strong contributions from higher-order
magnetic moments \citep{Kochukhov2011,Shultz2018} and complex light
curves resembling CPVs \citep{Mikulasek2020,Bouma2024}.  This suggests
that the photometric complexity of CPVs may be related to magnetic
fields with highly multipolar contributions.  \added{While this
suggestion is consistent with the wide range of axisymmetric and
non-axisymmetric topologies seen in mid- and late-M dwarfs
(e.g.~\citealt{Donati2006,Kochukhov2017,Shulyak2019,Bellotti2024}),
future work is needed to verify whether CPVs indeed select for
complex, non-axisymmetric fields.}

@@

\begin{acknowledgements}
  We thank B.~Tofflemire, A.~Weinberger, and L.~Hillenbrand for
  conversations that significantly informed this work; J.~Spake,
  J.~Winn, J.-F~Donati, and R.~Townsend for feedback on the
  manuscript; and A.~Howard and H.~Isaacson for their assistance
  reducing the HIRES spectra.

>>>>

% upon AAS submission
%\documentclass[11pt,twocolumn,tighten,linenumbers]{aastex7}
%\documentclass[11pt,twocolumn,tighten,linenumbers,trackchanges]{aastex7}
% drafting / arxiv
\documentclass[11pt,twocolumn,tighten]{aastex7}
%\turnoffedit

@@

\received{23 April 2025}
\revised{29 May 2025}
\accepted{6 June 2025}
\shorttitle{CPV Plasma Torus}

@@

\begin{figure*}[!t]
  \centering
  \includegraphics[width=0.925\textwidth]{f2.pdf}
  \vspace{-0.2cm}
  \caption{{\bf Photometric evolution of TIC\,141146667 around the Keck/HIRES
  observation (green bar).}  {\bf a,} TESS simple aperture photometry.
  Data gaps were caused by stray light from the Earth (BTJD 3356-3358.5) and
  Moon (BTJD 3359.5-3360.5).  Raw two minute data are in gray; black
  time-averages to ten minute sampling.  {\bf b-c,} Folded TESS light
  curve before and after spectroscopy.  Black now phase-averages to
  100 points per 3.93\ hour cycle.  During BTJD 3352-3356, a state
  switch occurred near BTJD 3353, and the dip at $\phi$$\approx$0.8
  \added{became less pronounced}.  While the \added{dip spanning
  $\phi$=0.6-1.1} was present both before and after the HIRES
  sequence, its photometric shape evolved during the data gap.  The
  orange bar denotes times of spectroscopic transits for the inner two
  H$\alpha$ clumps observed with HIRES (see Figure~\ref{fig:spec}).}
  \label{fig:fulllc}
\end{figure*}

\begin{figure*}[!t]
  \centering
  \includegraphics[width=0.99\textwidth]{f3.pdf}
  \vspace{-0.3cm}
  \captionsetup{labelformat=moviefmt,labelsep=colon}
	\caption{\textbf{Emission from plasma clumps orbiting
TIC\,141146667.}
  The online movie,
  \href{https://lgbouma.com/movies/TIC141146667_sixpanel.mp4}{{\bf
  available here}},
  shows the spectral evolution over five hours.
  {\bf a,} TESS light curve from 5 February 2024 to 26
  February 2024 folded on the 3.93\ hour period.  Black
  points are phase-averaged; gray are the raw data.
  {\bf b,} Keck/HIRES H$\alpha$ spectra from 17 February 2024.
  The continuum is set to unity, and the darkest color is at twice the
  continuum to accentuate emission outside the line core ($|v/v_{\rm
  eq}|>1$, for the equatorial velocity $v_{\rm eq}$=130\,\kms).  While
  emission within the line core comes from the star's chromosphere, the
  sinusoidal emission features are most readily described by a warped
  plasma torus.
  {\bf c,} Individual epochs of Panel b, visible in the online movie.
  The dotted line shows the time-averaged spectrum, $f_{\langle t
  \rangle}$.
  {\bf d,} As in Panel a, but overplotting the
  median-normalized H$\alpha$ light curve at $|v/v_{\rm eq}|<1$.
  {\bf e,} As in Panel b, after subtracting $f_{\langle t
  \rangle}$.  The line core shows H$\alpha$ excesses and decrements
  advancing from the blue to red wings.
  The asymmetric color stretch mirrors the dynamic range of the data.
  {\bf f,} Individual epochs of Panel e, visible in the online
  movie.  \added{A separate version with best-fit sinusoids from
  Appendix~\ref{subsec:halpha} overplotted is also
  \href{https://lgbouma.com/movies/TIC141146667_sixpanel_sinusoids.mp4}{{\bf
  available}}.}}
  \label{fig:spec}
\end{figure*}

@@

The other CPV analog is the $\sigma$~Ori~E variables, a rare subset of
B stars with radiatively-driven winds that accumulate into warped
plasma tori \citep{Townsend2005}.  These tori tend to have dense
antipodal accumulations of plasma sculpted by tilted-dipole magnetic
fields; these clumps produce broadband optical variability through
bound-free scattering \citep{Townsend2005} and Thomson scattering
\citep{Berry2022}.  For $\sigma$~Ori~E and most of its analogs, the
result is simple light curves that resemble eclipsing binaries, and
time-dynamic H$\alpha$ spectra similar to Figure~\ref{fig:spec}. 
\added{The few known exceptions---HD~37776, HD~64740, and HD~176582---}have complex
light curves resembling CPVs \citep{Mikulasek2020,Bouma2024} and
spectropolarimetric measurements that suggest non-dipolar field
contributions \citep{Kochukhov2011,Shultz2018}.  This suggests that
the photometric complexity of CPVs may be related to magnetic fields
with highly multipolar contributions.  \added{While this suggestion is
consistent with the wide range of axisymmetric and non-axisymmetric
topologies seen in mid- and late-M dwarfs
(e.g.~\citealt{Donati2006,Kochukhov2017,Shulyak2019,Bellotti2024}),
future work is needed to verify whether CPVs indeed select for
complex, non-axisymmetric fields.}

@@

\begin{acknowledgements}
  We thank B.~Tofflemire, A.~Weinberger, and L.~Hillenbrand for
  conversations that significantly informed this work; J.~Spake,
  J.~Winn, J.-F~Donati, R.~Townsend, and O.~Kochukhov for feedback on the
  manuscript; and A.~Howard and H.~Isaacson for their assistance
  reducing the HIRES spectra.
